\textbf{ Spectrophotometric parallaxes with linear models:\\ Accurate distances for luminous red-giant stars }

### 2019-07-29 paper/sp.tex
reference
response 2

There is a huge range of complexity or capacity in data-driven
models, from linear regressions (what we use here) to the most advanced machine-learning
methods (like deep neural networks).
In this work we stay on the low-complexity side of this, as
we do not want to face the (literally combinatoric) choices in model structure that
advanced methods bring.
And we want to control or understand the smoothness or flexibility of the model.
Deep networks can model extremely complex function spaces, which is good in some
contexts, but bad when you know that the functional dependencies you hope to
model are smooth:
In that case, much of the information in the data can get spent on learning the
smoothness of the function.

@@

Another validation of the results can be obtained by looking at known
clusters or spatially compact objects in the data.
In \figurename~\ref{fig:clusters}, we show the parallax distribution from \gaia\ and
the spectrophotometric-parallax distribution from our model for stars that are
astrometrically confirmed members of three of the stellar clusters in which we
validated our results.
The cluster members were found by hand, looking at sky and proper-motion distributions
centered on literature values.
In each case we are plotting only very securely identified cluster members; that
is, membership is conservative.
We chose the three displayed clusters
to span some range in metallicity and age and extinction, and therefore test
the accuracy of the method for different kinds of populations.
With the exception of \acronym{M71}, the clusters we tested
indicate that both \gaia\ and the spectrophotometric parallaxes appear to be unbiased
for these clusters and the range in abundances and ages they represent.
The case of cluster \acronym{M71} is troubling, but after searching for trends with
housekeeping data (also see \figurename~\ref{fig:residuals}) we don't find any correlations
with parallax offsets, although \acronym{M71} is a higher-extinction cluster.
\begin{figure}
~\hfill\includegraphics[width=0.7\textwidth]{clusters.pdf}\hfill~
\caption{The \gaia\ astrometric parallaxes $\gparallax_n$
  and spectrophotometric parallaxes $\sparallax_n$ for member stars for
  three of the stellar clusters in the \apogee\ observing footprint. The spectrophotometric
  parallaxes and their inverse-variance-weighted mean and the astrometric parallaxes
  and their inverse-variance-weighted mean
  are shown as vertical lines.
  Cluster members were identified to be close to the cluster in celestial coordinates
  and comoving in proper motion.\label{fig:clusters}}
\end{figure}

@@

It is also interesting to compare these estimates to other, previous
spectroscopic distance efforts.
We performed comparisons with two other efforts (\citealt{sch, que}).
In both cases, the agreement is better at smaller distances (larger
parallaxes) and worse at larger distances, with the distances
(parallaxes) presented here being smaller (larger) on average than the other
investigation results for the very most distant (and hence most difficult) stars.
In detail, the comparison with the more recent sample (\citealt{que})
are better:
The scatter is at the 15-percent level for stars at distances
$<8$\,kpc), and there is no visible bias between the results at these
distances.
The distances presented here also compare well with those from a recent
machine-learning approach (\citealt{leung}), which appeared while this
paper was under review.

@@

However, our choice of features is very rigid:
In principle the data themselves should tell us what features to include.
In that direction, it would make sense to choose features using a more complex
technique like deep learning or an auto-encoder.
These methods generate good features automatically.
They also involve an enormous set of hyper-parameter-like choices,
and they involve sacrificing certain kinds of interpretability.
All that said, we expect that a better set of features would do better; in this sense
the project here is just a first step towards precise spectrophotometric distance
estimates.

@@

The answer to the second question is that almost certainly we could have done better!
Since models like Gaussian processes and deep networks subsume the linear model
(at least approximately), they could have delivered better---or at least non-worse---results.
And indeed, after the the appearance of this manuscript, such an investigation
has appeared (\citealt{leung}) and it does produce very good results.
The issues with going to a much more complex model are manifold, however:
We would have had many more decisions to make and many more hyper-parameters
to set.
We would have used some (or maybe much) of the information in the data to learn
the simple fact that the problem is close to linear; that is, if you give a model
a lot of freedom or capacity, it has to use a lot of the data to learn what part
of that capacity it really needs.
A more complex model might also have edge issues: Very flexible models don't extrapolate
outside the convex hull of their training data, and our training data was not
balanced in terms of parallax and dust attenuation.
In the  linear model it is also---unlike in more flexible models---trivial to propagate
uncertainties from the feature space to the prediction.
We used that, above, to propagate uncertainty in \equationname~(\ref{eq:unc}).
And finally---in principle---linear models are better for visualization and interpretation;
the linear model contains essentially only first derivatives in the data space, which are
relatively easy to understand.

@@

For an example of how the model might be interpreted, we can look at
the photometric part of the linear model and see if the combination of
photometry does indeed look like a dust-corrected absolute magnitude,
as we expected (see \sectionname~\ref{sec:data}).
It does! Indeed the model automatically finds a photometric
combination that includes a term close to $H-W_2$, which is known as a
good extinction estimator.
And we find that on the two-fold A/B split, the photometric model is extremely stable.
As for the spectroscopic-pixel part of the model, we find that interpretation is
less simple, for two reasons:
One is that the number of pixels is enormous, so the model has huge flexibility.
Another is that although the two splits, A and B, both give similar-performing
models, they have (in detail) different linear coefficients in the spectral space.
That is, the model is not as stable in the spectroscopic domain as it is in the
photometric domain.
This is a generic feature of high-capacity machine-learning methods.
And it means that we did not obtrain as much insight from the spectroscopic model
as we had hoped up-front.

Although the model, training, and prediction make no use whatsoever of stellar
models, we did use stellar models indirectly to select the Parent Sample of stars:
We used the \apogee\ pipeline (\citealt{aspcap}) surface-gravity $\logg$ measurements.
This tiny use of stellar models could have been obviated by selecting stars not
on the basis of the derived physical parameter, but instead selecting stars to
be similar, observationally, in the spectral space.
That would have lead to a clean sample and made the project independent completely
of stellar models.

@@

Because of our train-and-test framework, each star is given a spectrophotometric
parallax based on coefficients derived from training on a complementary Training Set sample, of
which that star is not a member.
This framework ensures that the \gaia\ parallax $\gparallax_m$ of star $m$ is never used,
even indirectly, in generating the spectrophotometric parallax $\sparallax_m$ of star $m$.
That means that the parallax estimates generated here and given in \tablename~\ref{tab:output}
are statistically independent of the \gaia\ parallaxes and can be combined with them in
na\"ive ways, for example in inverse-variance-weighted averages.

>>>>

There is a huge range of complexity or capacity in data-driven
models, from linear regressions (what we use here) to the most advanced machine-learning
methods (like deep neural networks).
In this work we stay on the low-complexity side of this, as
we do not want to face the (literally combinatoric) choices in model structure that
advanced methods bring.
And we want to control or understand the smoothness or flexibility of the model.
Deep networks can model extremely complex function spaces, which is good in some
contexts, but bad when you know that the functional dependencies you hope to
model are smooth.
In that case, much of the information in the data can get spent on learning the
smoothness of the function.

@@

Another validation of the results can be obtained by looking at known
clusters or spatially compact objects in the data.
In \figurename~\ref{fig:clusters}, we show the parallax distribution from \gaia\ and
the spectrophotometric-parallax distribution from our model for stars that are
astrometrically confirmed members of three of the stellar clusters in which we
validated our results.
The cluster members were found by hand, looking at sky and proper-motion distributions
centered on literature values.
In each case we are plotting only very securely identified cluster members; that
is, membership is conservative.
We chose the three displayed clusters
to span some range in metallicity, age, and extinction, and therefore test
the accuracy of the method for different kinds of populations.
With the exception of \acronym{M71}, the clusters we tested
indicate that both \gaia\ and the spectrophotometric parallaxes appear to be unbiased
for these clusters and the range in abundances and ages they represent.
The case of cluster \acronym{M71} is troubling, but after searching for trends with
housekeeping data (also see \figurename~\ref{fig:residuals}) we don't find any correlations
with parallax offsets, although \acronym{M71} is a higher-extinction cluster.
\begin{figure}
~\hfill\includegraphics[width=0.7\textwidth]{clusters.pdf}\hfill~
\caption{The \gaia\ astrometric parallaxes $\gparallax_n$
  and spectrophotometric parallaxes $\sparallax_n$ for member stars for
  three of the stellar clusters in the \apogee\ observing footprint. The spectrophotometric
  parallaxes and their inverse-variance-weighted mean and the astrometric parallaxes
  and their inverse-variance-weighted mean
  are shown as vertical lines.
  Cluster members were identified to be close to the cluster in celestial coordinates
  and comoving in proper motion.\label{fig:clusters}}
\end{figure}

@@

It is also interesting to compare these estimates to other, previous
spectroscopic distance efforts.
We performed comparisons with two other efforts (\citealt{sch, que}).
In both cases, the agreement is better at smaller distances (larger
parallaxes) and worse at larger distances, with the distances
(parallaxes) presented here being smaller (larger) on average than the other
investigation results for the very most distant (and hence most difficult) stars.
In detail, the comparison with the more recent sample (\citealt{que})
is better:
The scatter is at the 15-percent level for stars at distances
$<8$\,kpc, and there is no visible bias between the results at these
distances.
The distances presented here also compare well with those from a recent
machine-learning approach (\citealt{leung}), which appeared while this
paper was under review.

@@

However, our choice of features is very rigid.
In principle the data themselves should tell us what features to include.
In that direction, it would make sense to choose features using a more complex
technique like deep learning or an auto-encoder.
These methods generate good features automatically.
They also involve an enormous set of hyper-parameter-like choices,
and they involve sacrificing certain kinds of interpretability.
All that said, we expect that a better set of features would do better; in this sense
the project here is just a first step towards precise spectrophotometric distance
estimates.

@@

The answer to the second question is that almost certainly we could have done better!
Since models like Gaussian processes and deep networks subsume the linear model
(at least approximately), they could have delivered better---or at least similar---results.
And indeed, after the the appearance of this manuscript, such an investigation
has appeared (\citealt{leung}) and it does produce very good results.
The issues with going to a much more complex model are manifold, however.
We would have had many more decisions to make and many more hyper-parameters
to set.
We would have used some (or maybe much) of the information in the data to learn
the simple fact that the problem is \emph{close} to linear.
A more complex model might also have edge issues, because flexible models don't extrapolate
outside the convex hull of their training data.
In the  linear model it is also---unlike in more flexible models---trivial to propagate
uncertainties from the feature space to the prediction.
We used that, above, to propagate uncertainty in \equationname~(\ref{eq:unc}).
And finally---in principle---linear models are better for visualization and interpretation;
the linear model contains essentially only first derivatives in the data space.

@@

For an example of how the model might be interpreted, we can look at
the photometric part of the linear model and see if the combination of
photometry does indeed look like a dust-corrected absolute magnitude,
as we expected (see \sectionname~\ref{sec:data}).
It does! Indeed the model automatically finds a photometric
combination that includes a term close to $H-W_2$, which is known as a
good extinction estimator.
And we find that on the two-fold A/B split, the photometric model is extremely stable.
As for the spectroscopic-pixel part of the model, we find that interpretation is
less simple, for two reasons:
One is that the number of pixels is enormous, so the model has huge flexibility.
Another is that although the two splits, A and B, both give similar-performing
models, they have (in detail) different linear coefficients in the spectral space.
That is, the model is not as stable in the spectroscopic domain as it is in the
photometric domain.
This is a generic feature of high-capacity machine-learning methods.
And it means that we did not obtain as much insight from the spectroscopic model
as we had hoped up-front.

Although the model, training, and prediction make no use whatsoever of stellar
models, we did use stellar models indirectly to select the Parent Sample of stars.
We used the \apogee\ pipeline (\citealt{aspcap}) surface-gravity $\logg$ measurements.
This tiny use of stellar models could have been obviated by selecting stars not
on the basis of the derived physical parameter, but instead selecting stars to
be similar, observationally, in the spectral space.
That would have lead to a clean sample and made the project independent completely
of stellar models.

@@

Because of our train-and-test framework, each star is given a spectrophotometric
parallax based on coefficients derived from training on a complementary Training Set sample, of
which that star is not a member.
As we emphasize above, this also ensures that the \gaia\ measurement $\gparallax_m$ of star $m$ is never used,
even indirectly, in generating the spectrophotometric estimate $\sparallax_m$ for that star $m$.
This, in turn, ensures statistical independence of the
parallax estimates generated here from the \gaia\ parallaxes.
The two could even be combined na\"ively in an inference.
